```{r global options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

# Lab 5: Homework Problems
Gosia Kukiełka

### Dependencies
Load necessery libraries.

```{r load}
library(devtools)
library(Biobase)
library(sva)
library(broom)
library(tidyverse)
library(data.table)
library(ggplot2)
library(gplots)
library(RColorBrewer)
library(gridExtra)
library(bladderbatch)
```
### Load data
```{r}
# sample info
data(bladderdata)
pheno = pData(bladderEset)
# expression data
edata = exprs(bladderEset)
```
### Prepare data
```{r}
#sumna <- apply(edata, 1, function(x) sum(is.na(x)))
row.variances <- apply(edata, 1, function(x) var(x))
edata <- edata[row.variances < 6,]
edata.log <- log2(edata)
#edata <- scale(edata, scale=FALSE, center=TRUE)
```
## Homework Problem 1:
Create a table to show the batch effects (refer to Figure 1 in Gilad and Mizrahi-Man, 2015). 
There are 5 batches (`pheno$batch`); how are biological variables and other variables related to study design are 
distributed among those 5 batches? Explain what could be a problem. Prepare this into a PDF file.
```{r}
pdf(file = "../figures/kukielka_problem1.pdf", width=18, height=8)

batch = as.factor(pheno$batch)
g1 <- ggplot() + geom_histogram(data=pheno, aes_string("cancer"), stat="count") + 
        facet_wrap(~batch, nrow=1) + ylab(label = "Cancer") + xlab(label=" ")

g2 <- ggplot() + geom_histogram(data=pheno, aes_string("outcome"), stat="count") + 
        facet_wrap(~batch, nrow=1) + ylab(label = "Outcome") + xlab(label=" ")

text = paste("\nThe main problem here is that the variables are not equally distributed among the batches and as a result, the data that we get can be 'contaminated' ",
              "\nwith patterns unrelated to the biological signal of our interest. The histogrammes show that different types of bladder cancers are separated quite well",
              "\nwhen it comes to batches (mTCC is mainly in the 1t batch, mTCC-CIS in the 2nd and mTCC+CIS in the 3rd) and so we can imagine a situation in which the five",
              "\nbatches correspond to the days in which data was gathered or different rearchers resposible for gathering it.",
             "\nSuch variables greatly influence the results of our analysis and, if left unacccounted for, can lead to false biological conclusions.")

t<- ggplot() + annotate("text", x = 4, y = 25, size=5, label = text) + theme_void()

grid.arrange(g1, g2, t, nrow = 3, top = "Batch")
dev.off()
```
## Homework Problem 3:
Compute the correlation statistics among columns. 
```{r}
corr_matrix <- cor(edata, method = 'pearson')
```
Visualize sample correlations using heatmap.2
```{r}
png("../figures/kukielka_problem3.png",height=700,width=700)

my_palette <- colorRampPalette(c("blue", "white", "darkred"))(n = 299)

heatmap.2(corr_matrix,
          main = "Bladder Cancer Data Sample Correlation", # heat map title
          notecol="black",      # change font color of cell labels to black
          density.info="none",  # turns off density plot inside color legend
          trace="none",         # turns off trace lines inside the heat map
          margins =c(12,9),     # widens margins around plot
          col=my_palette,       # use on color palette defined earlier 
          dendrogram="row",     # only draw a row dendrogram
          labCol = rev(pheno$outcome),
          labRow = pheno$cancer,
          xlab = "Outcome",
          ylab = "Cancer",
          revC = TRUE)

dev.off()
```